\chapter{Introduction}
\label{chap:introduction}

Organizations executing large-scale Capital Expenditure (CAPEX) portfolios face a fundamental
tension: multiple concurrent projects compete for the same finite pool of specialized engineering
skills, yet conventional planning tools offer no objective mechanism for resolving these conflicts
in a strategically coherent way. When resource shortfalls arise — as they inevitably do — portfolio
managers must decide which projects continue uninterrupted and which are slowed down or paused.
These decisions are today made largely through negotiation, intuition, and political influence
rather than through systematic, data-driven analysis.

The consequence is a persistent misalignment between strategic portfolio priorities and operational
resource deployment. High-priority projects stall because critical specialists are committed
elsewhere; lower-priority work consumes capacity that cannot be redirected in time. An AI-driven
prescriptive framework that can objectively recommend which projects to suspend — and when to
resume them — would allow organizations to actively govern this trade-off rather than merely react
to it.

% ─────────────────────────────────────────────────────────────────────────────
\section{Problem Statement}
\label{sec:problem_statement}

Current resource management in CAPEX portfolio environments is predominantly reactive and
subjective. Existing tools — spreadsheets, generic ERP modules, and project portfolio management
(PPM) software — can visualize resource conflicts after they materialize, but they cannot
prescribe which projects to pause or resume in order to maximize strategic value under binding
resource constraints. As a result:

\begin{itemize}
  \item Suspension and resumption decisions are made ad hoc, driven by seniority or urgency
        rather than strategic portfolio alignment.
  \item Specialized engineering skills are allocated to whichever project escalates first,
        not to whichever project delivers the highest strategic return.
  \item There is no systematic mechanism to evaluate the downstream impact of suspending a
        project — on cost, schedule, risk, and overall portfolio value.
\end{itemize}

This gap is particularly acute in capital-intensive industries (energy, infrastructure,
manufacturing) where CAPEX programs span multiple years, involve dozens of concurrent projects,
and depend on a narrow set of scarce specialist roles (e.g., systems integrators, project
controls engineers, domain architects). The inability to objectively arbitrate competing resource
claims causes schedule slippage, budget overruns, and erosion of strategic intent.

Advances in prescriptive AI — combining optimization, simulation, and priority-scoring models —
offer a principled foundation for automating these trade-off decisions \citep{gartner2024hype}.
The AI-4-SME Framework \citep{peter2025ai4sme}, developed at FHNW School of Business, provides a
structured methodology for designing and deploying such AI solutions in enterprise contexts.

% ─────────────────────────────────────────────────────────────────────────────
\section{Thesis Statement, Research Questions and Objectives}
\label{sec:thesis_rq_objectives}

\subsection*{Thesis Statement}

This thesis investigates how an AI-driven prescriptive framework can be designed to support
strategic resource governance in CAPEX portfolio management. The core mechanism under study is
the generation of objective suspension-and-resumption recommendations: the system temporarily
pauses lower-priority projects to free specialized engineering skills for projects of higher
strategic importance, thereby preserving strategic alignment under binding resource constraints.

The thesis statement is:
\begin{quote}
  \textit{The implementation of an AI-driven prescriptive framework secures the strategic
  alignment of a CAPEX portfolio under resource constraints by generating objective
  `suspension-and-resumption' recommendations. This process optimizes project execution by
  temporarily pausing lower-priority tasks to release specialized engineering skills for
  projects of higher strategic importance.}
\end{quote}

\subsection*{Research Questions}

\noindent\textbf{Main Research Question:}
\begin{quote}
  How can an AI-supported prescriptive framework enhance the strategic alignment of a CAPEX
  portfolio by automating project de-prioritization recommendations during specialized skill
  shortages?
\end{quote}

\noindent\textbf{Sub-Questions:}

\begin{description}
  \item[SQ1] How can the temporal demand for specialized technical skill pools be accurately
  modeled across a portfolio of 300+ ongoing projects?

  \item[SQ2] How can a dynamic weighting framework (Multi-Criteria Decision Making) be
  designed to integrate diverse and evolving importance indicators into an automated
  de-prioritization logic?

  \item[SQ3] What is the impact of systematic project suspension and resumption on overall
  resource availability and the successful execution of high-priority initiatives?
\end{description}

\subsection*{Research Objectives}

\begin{enumerate}
  \item Develop a strategic priority scoring model that quantifies the relative importance of
        projects within a CAPEX portfolio across multiple dimensions (financial, strategic,
        risk, time-criticality).
  \item Design a prescriptive AI framework that combines priority scoring with resource
        constraint modeling to generate suspension-and-resumption recommendations.
  \item Implement a prototype of the framework and validate its recommendations against expert
        judgment and simulated portfolio scenarios.
  \item Derive generalizable design principles for AI-driven prescriptive systems in
        resource-constrained portfolio management contexts.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Delineations and Limitations}
\label{sec:delineations}

\textbf{Delineations:}
\begin{itemize}
  \item The research focuses on CAPEX portfolio management in medium-to-large organizations;
        micro-enterprises and purely operational (OPEX) contexts are excluded.
  \item The prescriptive framework addresses \textit{suspension-and-resumption} decisions at
        the project level; intra-project task scheduling (CPM/PERT) is outside scope.
  \item The resource constraints considered are \textit{specialized human skills}; financial,
        equipment, and infrastructure constraints are not modeled.
  \item Empirical validation uses anonymized data from a partner organization or synthetic
        portfolio scenarios; multi-industry generalization is a direction for future work.
\end{itemize}

\textbf{Limitations:}
\begin{itemize}
  \item The quality of recommendations depends on the accuracy and completeness of project
        priority data, which may be politically sensitive or inconsistently defined in practice.
  \item The study covers a defined research period; long-term organizational adoption and
        behavioral change are not assessed.
  \item The framework optimizes for strategic alignment as modeled; unmeasured organizational
        factors (politics, client commitments) may override recommendations in practice.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Assumptions}
\label{sec:assumptions}

The following assumptions are taken as true for the purpose of this study:

\begin{itemize}
  \item Organizations can define and quantify the strategic importance of their CAPEX projects,
        even if the scoring model requires elicitation and calibration.
  \item Project suspension and resumption are organizationally feasible actions — i.e., there
        exist no absolute contractual or regulatory prohibitions preventing temporary pauses.
  \item Specialized engineering skills are the binding resource constraint; other resources
        (budget, equipment) are assumed not to be simultaneously binding.
  \item Stakeholders and subject-matter experts who provide priority scores and validation
        feedback respond honestly and to the best of their knowledge.
  \item The AI-4-SME Framework \citep{peter2025ai4sme}, developed at FHNW, provides a
        methodologically sound structure for the Design, Build, and Run phases of this research.
  \item The regulatory environment (GDPR, EU AI Act) remains stable over the research period.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Significance}
\label{sec:significance}

\textbf{Theoretical significance:}
This study contributes to the body of knowledge in Business Information Systems by introducing a
prescriptive AI approach to the problem of strategic resource governance in CAPEX portfolios. It
extends existing work on resource-constrained project scheduling by incorporating strategic
priority scoring as a first-class input to optimization, and by framing suspension-and-resumption
as a deliberate governance mechanism rather than an exceptional response. The research also
advances design science knowledge on how AI prescriptive systems can be constructed to produce
explainable, actionable recommendations in high-stakes organizational settings.

\textbf{Practical significance:}
Portfolio managers in capital-intensive organizations gain a systematic, objective basis for
making suspension-and-resumption decisions — reducing reliance on ad hoc negotiation and
political influence. The framework enables proactive reallocation of scarce engineering talent
to projects where it creates the most strategic value, directly reducing schedule slippage,
budget overruns, and misalignment between investment intent and execution reality. Integration
with existing ERP and PPM systems provides a pathway to operational adoption.

% ─────────────────────────────────────────────────────────────────────────────
\section{Chapter Map}
\label{sec:chapter_map}

\textbf{Chapter 1 — Introduction:}
Introduces the research problem, background, thesis statement, research questions and objectives,
delineations and limitations, assumptions, and significance of the study.

\textbf{Chapter 2 — Literature Review:}
Reviews existing work on CAPEX portfolio management, resource-constrained project scheduling,
strategic priority scoring, and prescriptive AI. Identifies the research gap that motivates
this thesis.

\textbf{Chapter 3 — Research Method:}
Describes the research design, philosophical stance, methodological choices, data collection
and analysis strategy, and work plan including milestones, risks, and dependencies.

\textbf{Chapter 4 — Conclusion:}
Summarizes anticipated theoretical and practical contributions and outlines directions for
future research.

\vspace{1em}
\noindent The graphical chapter map below illustrates the logical flow from the research questions
to the thesis chapters:

\vspace{1em}
\begin{center}
\begin{tikzpicture}[
  box/.style={draw, rounded corners, minimum width=3.8cm, minimum height=0.8cm,
              text centered, font=\small},
  rq/.style={draw, rounded corners, minimum width=3.5cm, minimum height=0.75cm,
             text centered, font=\footnotesize, fill=gray!15},
  arr/.style={->, thick}
]
  % Chapter 1
  \node[box, fill=blue!10] (ch1) at (0,0) {Chapter 1: Introduction};

  % SQ nodes
  \node[rq] (sq1) at (-4.5,-2)  {SQ1: Skill demand modelling};
  \node[rq] (sq2) at (0,-2)     {SQ2: MCDM weighting};
  \node[rq] (sq3) at (4.5,-2)   {SQ3: Suspension impact};

  \draw[arr] (ch1) -- (sq1);
  \draw[arr] (ch1) -- (sq2);
  \draw[arr] (ch1) -- (sq3);

  % Chapter 2
  \node[box, fill=green!10] (ch2) at (-3,-4) {Chapter 2: Literature Review};
  \draw[arr] (sq1) -- (ch2);
  \draw[arr] (sq2) -- (ch2);

  % Chapter 3
  \node[box, fill=orange!10] (ch3) at (3,-4) {Chapter 3: Research Method};
  \draw[arr] (sq2) -- (ch3);
  \draw[arr] (sq3) -- (ch3);

  % Conclusion
  \node[box, fill=red!10] (ch4) at (0,-6) {Chapter 4: Conclusion};
  \draw[arr] (ch2) -- (ch4);
  \draw[arr] (ch3) -- (ch4);
\end{tikzpicture}
\end{center}
