\chapter{Literature Review}
\label{chap:literature_review}

The literature review maps the body of knowledge relevant to the research problem and
systematically identifies the gap that motivates this thesis. Following a structured search
strategy, the review synthesises eight thematic clusters: (A)~CAPEX portfolio management and
strategic alignment, (B)~project preemption and suspension mechanisms, (C)~resource-constrained
project scheduling with skill considerations, (D)~prescriptive analytics and AI-driven decision
support, (E)~multi-criteria decision making for project prioritisation, (F)~workforce and skill
demand forecasting, (G)~design science research, and (H)~hybrid AI combining symbolic and
subsymbolic methods including the AI-4-SME framework. The chapter closes with a cross-cluster gap
analysis that directly motivates the thesis research questions.

% ─────────────────────────────────────────────────────────────────────────────
\section{Search Strategy}
\label{sec:search_strategy}

The systematic search was conducted on \textbf{Scopus} as the primary academic database,
supplemented by targeted searches on IEEE Xplore, ACM Digital Library, and INFORMS PubsOnline
for conference and journal papers not fully indexed at the time of search. The search applied
Boolean keyword combinations across eight thematic clusters, as summarised in
Table~\ref{tab:keywords}. A two-stage screening procedure was applied: title and abstract
screening followed by full-text relevance assessment against the four research questions. Only
papers rated Medium or High relevance were retained for synthesis. In total, 35 papers were
screened and included, of which 33 were subject to direct PDF analysis; key findings below are
grounded in full-text reading.

\begin{table}[htbp]
  \centering
  \caption{Keyword clusters used in the systematic Scopus search}
  \label{tab:keywords}
  \small
  \begin{tabular}{@{}lp{10.5cm}@{}}
    \toprule
    \textbf{Cluster} & \textbf{Search Terms} \\
    \midrule
    A — Portfolio & \texttt{CAPEX portfolio management}, \texttt{capital expenditure
      portfolio}, \texttt{project portfolio strategic alignment},
      \texttt{portfolio value optimisation} \\[0.3em]
    B — Preemption & \texttt{project preemption}, \texttt{preemptive RCPSP},
      \texttt{activity splitting}, \texttt{project suspension resumption} \\[0.3em]
    C — RCPSP & \texttt{resource-constrained project scheduling},
      \texttt{multi-skill scheduling}, \texttt{multi-project scheduling},
      \texttt{hierarchical skill scheduling} \\[0.3em]
    D — Prescriptive & \texttt{prescriptive analytics}, \texttt{decision support
      resource allocation}, \texttt{from predictive to prescriptive},
      \texttt{prescriptive analytics system} \\[0.3em]
    E — MCDM & \texttt{MCDM project prioritisation}, \texttt{AHP TOPSIS portfolio},
      \texttt{dynamic weighting strategic alignment}, \texttt{fuzzy AHP portfolio} \\[0.3em]
    F — Forecasting & \texttt{workforce skill demand forecasting},
      \texttt{temporal skill modelling}, \texttt{labour demand machine learning},
      \texttt{LSTM skill forecast} \\[0.3em]
    G — DSR & \texttt{Design Science Research methodology},
      \texttt{DSR information systems}, \texttt{Hevner design science} \\[0.3em]
    H — Hybrid AI & \texttt{hybrid AI knowledge graph ontology},
      \texttt{neuro-symbolic AI decision support}, \texttt{ESCO skill ontology},
      \texttt{composite AI enterprise} \\
    \bottomrule
  \end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{CAPEX Portfolio Management and Strategic Alignment}
\label{sec:lit_capex}

Project portfolio management (PPM) has evolved from a purely financial discipline into a
strategic governance function concerned with selecting, prioritising, and terminating projects
in ways that maximise organisational value. In a systematic literature review spanning seven
decades and 669 articles, \citet{hansen2022seven} map PPM research into seven overarching
streams and identify \textit{resource allocation under competing constraints} and
\textit{dynamic re-alignment to shifting strategic objectives} as the two most persistently
unresolved challenges in the field. A follow-up study by \citet{hansen2023principles} derives
seventeen normative PPM principles through a four-wave Lichtenberg analysis of 100 core
articles; their fourth-wave principles (P14--P17), which address readiness to change and
portfolio paradoxes, provide normative criteria for evaluating whether automated suspension
recommendations enhance or erode strategic coherence.

At the level of formal optimisation models, \citet{chiang2013strategic} propose a
multi-objective quadratic assignment formulation for IT project portfolios that integrates a
strategic alignment score alongside expected financial benefits, development cost, and
cross-project synergies~$f_{ij}$. Their Pareto-optimal solutions, obtained via evolutionary
algorithms, demonstrate that explicit strategic scoring produces portfolios that genuinely
reflect organisational priorities, whereas purely financial proxies lead to systematic
misalignment. Infrastructure dependencies~$d_{ii'}$ between projects further show that the
value of a portfolio cannot be assessed by summing individual project values in isolation.

\citet{rodriguezgarcia2025ai} extend this direction with a contemporary hybrid machine
learning and mathematical programming approach. Using XGBoost to predict synergy-adjusted
returns within a Desirability-Sustainability-Feasibility (DSF) framework and Gurobi for
exact optimisation, they demonstrate that synergy-aware portfolio selection yields 5--7\%
higher returns compared to static, synergy-agnostic baselines (ANOVA: $p < 0.0001$). The
approach scales to 500 projects in 64 seconds on commodity hardware, establishing
computational feasibility at real-world CAPEX portfolio scale. Crucially, however, the model
treats projects as binary selections without modelling temporal dynamics or project
suspension — a gap directly addressed by the present thesis.

\citet{rodriguescoelho2025strategic} provide empirical grounding for the strategic alignment
objective using Structural Equation Modelling with Partial Least Squares ($n = 198$
portfolio management practitioners) and Necessary Condition Analysis. Strategic alignment
significantly mediates the relationship between PPM and portfolio value (H1: $\Gamma = 0.566$,
$t = 5.573$, $p < 0.001$; H2: $\Gamma = 0.754$, $p < 0.001$). NCA further reveals that a
minimum 14.7\% performance level in risk and change management is a \textit{necessary}
condition for achieving a 30\% improvement in portfolio value — a threshold-based insight
with direct implications for the design of the MCDM scoring model.

% ─────────────────────────────────────────────────────────────────────────────
\section{Project Preemption and Suspension Mechanisms}
\label{sec:lit_preemption}

The concept of temporarily suspending an ongoing project activity to release a resource for
a higher-priority demand has a formal treatment in the operations research literature under
the label of \textit{preemption}. The foundational study by \citet{ballestin2008preemption}
demonstrates that allowing even a single preemptive interruption within a resource-constrained
project scheduling problem (RCPSP) yields statistically significant makespan reductions
relative to purely non-preemptive schedules. This result establishes preemption as an
operationally meaningful mechanism, not merely a theoretical relaxation.

\citet{vanpeteghem2010genetic} extend this analysis to multi-mode RCPSP, showing that a
genetic algorithm capable of selecting between preemptive and non-preemptive execution modes
consistently produces superior schedules when resource availability is uneven across the
planning horizon. Specifically, four execution modes per activity allow modelling of
part-time specialist engagement — a characteristic common to CAPEX programmes where engineers
are shared across multiple concurrent projects.

The most practically relevant model for the present thesis is provided by
\citet{hatamimoghaddam2024robust}, who introduce a bi-objective MINLP formulation for a
robust multi-skill, multi-mode RCPSP with \textit{partial preemption}. An activity is
suspended precisely when a required skill set becomes unavailable within a specified time
window and is resumed when the skill pool is replenished; NSGA-II is used to identify
Pareto-optimal trade-offs between makespan compression and disruption cost. A robust scenario
set accounts for demand uncertainty. This mechanism maps directly onto the
suspension-and-resumption logic proposed in the thesis, with specialised engineering skill
shortages serving as the trigger condition.

\citet{vanhoucke2025impact} contribute a bi-criteria analysis examining the trade-off between
makespan reduction and the number of interruptions induced by preemption across 480 problem
instances with time-varying resources. Their key empirical finding is that the
\textit{optimal suspension frequency is approximately three preemptions per project} before
disruption costs outweigh schedule benefits, and that this threshold varies with resource
availability patterns. This result directly informs the policy design of the proposed
prescriptive framework: a data-driven, dynamic policy is preferable to a fixed threshold rule.

Collectively, this body of work establishes that project suspension is a formally sound and
empirically effective mechanism for releasing constrained resources. However, all existing
models operate on single projects or small portfolios and employ rule-based or combinatorial
triggers rather than strategic priority scoring. The integration of strategic priority
intelligence into the preemption decision constitutes a first dimension of the research gap.

% ─────────────────────────────────────────────────────────────────────────────
\section{Resource-Constrained Project Scheduling with Skill Considerations}
\label{sec:lit_rcpsp}

The resource-constrained project scheduling problem (RCPSP) is one of the most extensively
studied problems in operations research. \citet{artigues2025fifty} trace fifty years of RCPSP
research in a comprehensive review, documenting the progression from exact branch-and-bound
methods to metaheuristics, and from single-project formulations to multi-skill, multi-project
variants. The review explicitly states that \textit{``machine learning for RCPSP is in its
infancy''} and identifies the combination of multi-skill resources, preemption, and multi-project
context as the most underexplored variant — a direct citation anchor for the research gap of
this thesis. \citet{hartmann2022updated} provide a complementary update across 230+ RCPSP
variants, noting that \textit{``the preemptive multi-skill variant is largely unaddressed''} —
further substantiating the gap.

For portfolios of the scale addressed in this thesis ($300+$ concurrent projects),
\citet{gomez2023survey} survey 120 resource-constrained multi-project scheduling problem
(RCMPSP) papers and demonstrate that shared resource pools across concurrent projects are
the primary source of scheduling conflicts. They explicitly note that the combination of
multi-skill resources, preemption, and multi-project context is absent from the existing
literature — directly mirroring the CAPEX environment of this research.

The modelling of human skill heterogeneity has received increasing attention.
\citet{snauwaert2021algorithm} introduce \textit{skill breadth} (number of distinct skills)
and \textit{skill depth} (proficiency level per skill) as two independent dimensions of the
multi-skilled RCPSP (MSRCPSP), demonstrating that bi-dimensional modelling improves
assignment quality by 15--20\% compared to single-dimension approaches. This distinction
mirrors the tiered engineering competency pools of large CAPEX programmes, where a domain
architect and a junior systems engineer both possess a skill but at different proficiency
levels. \citet{snauwaert2023classification} formalise a taxonomy of eight MSRCPSP variants
across four dimensions (skill substitution, hierarchical skills, learning effects, and
time-windows) and release standardised benchmark instances — providing the empirical
infrastructure for rigorous evaluation of the thesis's skill demand modelling approach.
Their most recent work \citep{snauwaert2025hierarchical} introduces six hierarchical-skill
MSRCPSP variants where a superior skill can substitute for an inferior one (but not vice
versa), directly mirroring tiered engineering competency structures.

The industrial feasibility of solving large-scale MSRCMPSP instances is demonstrated by
\citet{torba2024reallife}, who address a real-world scheduling problem at SNCF heavy
maintenance involving thousands of activities, six skill types, and shared skill pools across
concurrent projects. Their memetic algorithm combining evolutionary search with exact repair
procedures is the closest real-world analogue to the computational challenge of the present
thesis. \citet{haroune2023multiproject} address multi-project assignment under shared
employee skill pools with efficiency levels, minimising weighted tardiness — a formulation
that naturally accommodates project priority weights, establishing a modelling connection
between scheduling (Cluster~C) and MCDM priority scoring (Cluster~E).

% ─────────────────────────────────────────────────────────────────────────────
\section{Prescriptive Analytics and AI-Driven Decision Support}
\label{sec:lit_prescriptive}

Prescriptive analytics denotes the class of analytical methods that go beyond description
and prediction to generate actionable recommendations — explicit decisions or interventions
designed to achieve a specified objective under identified constraints. In a systematic
review of 56 key papers, \citet{lepenioti2020prescriptive} identify resource allocation as
the primary application domain (62\% of reviewed papers) and propose a four-layer taxonomy:
descriptive, diagnostic, predictive, and prescriptive. They emphasise that a closed-loop
feedback mechanism — through which recommendation outcomes inform model updates — is the
distinguishing architectural feature of mature prescriptive analytics systems.
\citet{frazzetto2019prescriptive} complement this from a data management perspective,
identifying constraint satisfaction, mathematical optimisation, and simulation as the three
core architectural pillars of any prescriptive analytics framework — a decomposition that
maps directly onto the three design components of the framework proposed in this thesis.

The theoretical foundation for embedding prediction within optimisation is provided by
\citet{bertsimas2020predictive}, who introduce the \textit{coefficient of prescriptiveness}
$\rho \in [0,1]$ to measure the decision value added by incorporating predictions into
optimisation. Using Sample Average Approximation (SAA) and Weighted Robust Optimisation
(WRO), they demonstrate 15--20\% cost reduction relative to purely predictive approaches
across several operational contexts. Their framework is particularly relevant to the
analytics chain of this thesis: temporal skill demand forecasts constitute the predictive
layer, whose output is the principal input to the suspension-and-resumption optimisation.

\citet{wissuchek2025prescriptive} provide the most recent and comprehensive IS-centric
treatment of prescriptive analytics systems (PAS) through a systematic literature review
of 52 papers. Their analysis identifies 23 architectural components spanning data pipeline,
predictive model, intervention/action selection module, explainability layer, and feedback
loop. Critically, the review finds that \textit{infrastructure CAPEX has only one existing
PAS case study} and that multi-project portfolio management with dynamic constraint sets is
explicitly identified as an open research gap — a direct empirical motivation for this thesis.
\citet{misic2020analytics} further contextualise prescriptive methods within the operations
management literature, noting that their comparative advantage over predictive methods is
most pronounced precisely when decision windows are short and resource constraints are binding
— both conditions that characterise CAPEX resource governance.

% ─────────────────────────────────────────────────────────────────────────────
\section{Multi-Criteria Decision Making for Project Prioritisation}
\label{sec:lit_mcdm}

Project prioritisation in the context of competing resource constraints is inherently a
multi-criteria problem: strategic importance, financial return, regulatory urgency,
stakeholder commitment, and execution risk are incommensurable attributes that must be
integrated into a coherent ranking or weighting scheme. \citet{kandakoglu2024mcdm} conduct
a systematic literature review of 127 papers applying MCDM methods to project portfolio
selection — covering AHP, PROMETHEE, TOPSIS, and DEA among others. Their central finding is
that \textit{dynamic re-prioritisation under constraint changes} is an open research gap:
existing MCDM applications to PPM are almost uniformly static, producing a priority ordering
at a single point in time rather than continuously updating weights as resource availability
and strategic context evolve. This constitutes a second major dimension of the research gap
of this thesis.

\citet{chatterjee2018fuzzy} address linguistic uncertainty in expert prioritisation through
Fuzzy AHP with triangular membership functions applied across five project evaluation
criteria (strategic fit, financial return, risk, resource availability, and time-criticality).
Their empirical demonstration that fuzzy weighting reduces rank reversal by 30\% compared to
crisp AHP is methodologically important: criteria weights assigned by portfolio managers are
inherently imprecise and may change over time, requiring a weighting framework that
accommodates vagueness without collapsing to arbitrary numerical scores. The dynamic
extension of this approach — where fuzzy weights are recalibrated in response to changing
resource conditions and project progress — forms the conceptual basis of Sub-Question~2.

The application of MCDM to capital-intensive industrial portfolios is examined by
\citet{alhashimi2017mcdm}, who compare AHP, PROMETHEE~II, and TOPSIS on an oilfield CAPEX
portfolio involving 12 projects with a combined investment of \$2.4~billion. The key finding
— that AHP and PROMETHEE show greater than 80\% rank coherence while TOPSIS diverges under
expert disagreement — underscores the need for systematic weight elicitation and sensitivity
analysis in any automated de-prioritisation system.

% ─────────────────────────────────────────────────────────────────────────────
\section{Workforce and Skill Demand Forecasting}
\label{sec:lit_forecasting}

Accurately modelling the temporal demand for specialised skill pools is a prerequisite for
any anticipatory suspension-and-resumption mechanism. Without a reliable forecast of when
and where skill shortages will materialise, suspension recommendations can only be generated
reactively — after conflicts have already disrupted project execution.

\citet{safarishahrbijari2018workforce} provide a systematic review of 87 workforce
forecasting papers, organising existing models into four families: exponential smoothing (best
for stable demand), ARIMA (seasonal patterns), Markov chain (state transitions between skill
pools), and system dynamics (feedback loops between supply and demand). Their key finding is
that hybrid model-switching approaches outperform single-family models in non-stationary
environments — a recommendation directly applicable to CAPEX portfolios where project
lifecycle phases create non-stationary skill demand patterns.

At the intersection of deep learning and skill demand forecasting, \citet{macedo2022skills}
apply RNN/LSTM models with representation learning to ten years of monthly LinkedIn skill
demand data. Multivariate models capturing inter-skill correlations outperform univariate
LSTM baselines by 18\% in multi-step forecast accuracy. Their approach explicitly handles
concept drift and the emergence of new skills — both phenomena relevant to CAPEX portfolios
operating across multi-year horizons with evolving engineering disciplines.

Temporal Fusion Transformers \citep{lim2021tft} represent the current state of the art for
multi-horizon time series forecasting with mixed-frequency inputs and static covariates. The
TFT architecture combines gating mechanisms and variable selection networks for feature
filtering with multi-head interpretable attention for long-range dependency modelling, and
quantile regression (P10/P50/P90) for probabilistic uncertainty bounds. On benchmark datasets
spanning electricity, traffic, retail, and financial volatility, TFT achieves 7--25\%
improvement in P50 loss and 3--8\% in P90 loss over LSTM, DeepAR, and ConvTrans baselines.
The architecture is a natural candidate for skill demand forecasting in CAPEX portfolios,
where project metadata (type, phase, contractual milestones) constitutes structured static
covariates modulating temporal demand patterns.

% ─────────────────────────────────────────────────────────────────────────────
\section{Design Science Research Methodology}
\label{sec:lit_dsr}

The thesis adopts Design Science Research (DSR) as its overarching methodological paradigm.
\citet{hevner2004design} establish the foundational framework for DSR in information systems,
prescribing seven guidelines for rigorous artifact design: (1)~design as an artifact,
(2)~problem relevance, (3)~design evaluation, (4)~research contributions, (5)~research rigor,
(6)~design as a search process, and (7)~communication of results. Their three-cycle model
— relevance cycle (environment), design cycle (IS research), and rigor cycle (knowledge base)
— provides the structural template for positioning the proposed prescriptive framework as a
valid IS research contribution. With over 10{,}000 citations, this paper constitutes the
primary methodological anchor of the thesis.

\citet{peffers2007dsrm} operationalise DSR into a six-phase process model: (1)~problem
identification and motivation, (2)~definition of objectives for a solution, (3)~design and
development, (4)~demonstration, (5)~evaluation, and (6)~communication. The model explicitly
supports multiple entry points (problem-centred, solution-centred, design-centred, and
contextual), accommodating the thesis's dual entry via both a practitioner problem and an
existing AI-4-SME methodological framework. The DSRM process model governs the sequence of
the thesis chapters: Chapters 1--2 correspond to phases 1--2; Chapter 3 to phases 3--5;
and Chapter 4 to phase 6.

% ─────────────────────────────────────────────────────────────────────────────
\section{Hybrid AI: Symbolic, Subsymbolic, and the AI-4-SME Framework}
\label{sec:lit_hybrid_ai}

A recurring limitation of purely data-driven approaches in enterprise decision support is
their dependence on the statistical regularity of historical data. In domains characterised
by low data volume, high structural complexity, or rapidly changing operational contexts —
all of which apply to CAPEX portfolio management — subsymbolic models alone are insufficient.
The paradigm of \textit{hybrid AI} addresses this by combining the complementary strengths
of symbolic and subsymbolic methods.

\textbf{Symbolic AI} encompasses methods that operate on explicit, human-interpretable
representations: ontologies, rule systems, logic programs, and knowledge graphs. Symbolic
methods are inherently interpretable, capable of encoding domain constraints formally, and
effective in low-data regimes. Their limitation is brittleness: they cannot generalise
beyond explicitly encoded knowledge.

\textbf{Subsymbolic AI} encompasses statistical learning methods — classical machine
learning, deep learning, and generative AI — that extract patterns from data without
explicit symbolic encoding. These methods are powerful pattern recognisers that generalise
to unseen inputs but lack interpretability, require large training sets, and cannot
natively enforce hard domain constraints.

\textbf{Hybrid AI} combines both paradigms. \citet{garcez2022neurosymbolic} formalise this
integration as \textit{neurosymbolic AI}, proposing a taxonomy of six system types along a
spectrum from loose coupling (Type~2: neural and symbolic components run independently) to
tight integration (Type~6: Logic Tensor Networks with symbolic constraints embedded in
neural loss functions). They identify gradient-based optimisation, modularity, and symbolic
language with constraint satisfaction as the three essential ingredients of robust
neurosymbolic systems. Their empirical evidence shows that neurosymbolic systems are
simultaneously more accurate and more interpretable than either component in isolation.

The Gartner Hype Cycle \citep{gartner2024hype} positions \textit{Composite AI} — the
enterprise term for hybrid AI that combines rule-based and ML approaches — at a 2--5 year
mainstream adoption horizon, with Responsible AI governance emerging as a business
imperative. The report identifies data strategy (data governance, knowledge graphs, quality
management) as the critical enabling prerequisite for AI deployment in enterprise contexts.

\textbf{The ESCO Ontology} provides the symbolic backbone for skill representation in this
thesis. Developed by the European Commission, ESCO is a multilingual, hierarchical taxonomy
of 13{,}900+ skills and competences formally linked to occupations and qualifications through
a structured ontological hierarchy with SKOS/OWL relationships
(\texttt{skos:broader}, \texttt{skos:narrower}, \texttt{skos:related})
\citep{le2014esco}. ESCO enables: (i)~consistent skill labelling across heterogeneous
project data sources; (ii)~ontological inference of skill substitutability
(i.e., which alternative skills can fulfil a demand when the primary skill pool is
unavailable); and (iii)~semantic enrichment of subsymbolic demand forecasting models through
ontology embeddings.

\textbf{The AI-4-SME Framework} \citep{peter2025ai4sme}, developed at FHNW School of
Business, provides the overarching methodological structure for translating the hybrid AI
architecture into an implementable solution. The framework's three-phase
Design--Build--Run methodology operates across three organisational levels (company,
process, task). Phase~1 (Design) uses BCG portfolio matrix analysis and capability mapping
to identify AI opportunities; a key analytical distinction is made between
Knowledge-Intensive Activities (KIAs) — requiring contextual judgement, problem solving, and
expert knowledge — and Data-Intensive Activities (DIAs) — requiring high-volume data
processing. In the CAPEX resource management context, strategic priority scoring and
suspension decision rationale generation are KIAs; temporal demand forecasting and constraint
propagation are DIAs. Phase~2 (Build) supports three build paths: data-based (TensorFlow,
PyTorch), knowledge-based (Neo4j, GraphDB, Stardog), or hybrid — the path directly
applicable to this thesis. Phase~3 (Run) addresses enterprise governance (GDPR, EU AI Act),
IT infrastructure integration (ERP/CRM APIs), and organisational change management. The
companion practice guide \citep{fhnw2024kmu} provides step-by-step workshop instructions and
worked case studies from Swiss companies implementing AI-4-SME solutions.

% ─────────────────────────────────────────────────────────────────────────────
\section{Research Gap and Positioning}
\label{sec:lit_gap}

The synthesis of the eight clusters reveals a coherent set of gaps at the intersection of the
reviewed domains. Table~\ref{tab:gap} summarises the state of knowledge per dimension and
identifies the contribution of the present thesis.

\begin{table}[htbp]
  \centering
  \caption{Research gap summary across literature clusters}
  \label{tab:gap}
  \small
  \begin{tabular}{@{}p{3.8cm}p{5.8cm}p{4.2cm}@{}}
    \toprule
    \textbf{Dimension} & \textbf{State of the Art} & \textbf{Gap / Thesis Contribution} \\
    \midrule
    Strategic alignment in PPM &
    Static portfolio selection with strategic scoring; synergy-aware AI optimisation
    \citep{chiang2013strategic, rodriguezgarcia2025ai} &
    No dynamic, AI-driven re-alignment triggered by real-time skill shortages \\[0.4em]

    Preemptive project scheduling &
    Rule-based and combinatorial preemption; bi-criteria optimal frequency~$\approx 3$
    \citep{hatamimoghaddam2024robust, vanhoucke2025impact} &
    No prescriptive AI layer translating portfolio-level strategic priorities into
    suspension triggers \\[0.4em]

    Multi-skill RCPSP at scale &
    Metaheuristics for industrial instances; hierarchical skill structures
    \citep{torba2024reallife, snauwaert2025hierarchical} &
    No integration with strategic MCDM weighting or AI-driven priority scoring;
    gap explicitly confirmed \citep{artigues2025fifty, hartmann2022updated} \\[0.4em]

    Prescriptive analytics systems &
    23-component PAS architecture; resource allocation as primary domain
    \citep{lepenioti2020prescriptive, wissuchek2025prescriptive} &
    Zero PAS case studies for multi-project CAPEX suspension-and-resumption;
    identified as open gap \citep{wissuchek2025prescriptive} \\[0.4em]

    MCDM for project prioritisation &
    Static, point-in-time ranking; 127 reviewed papers confirm no dynamic variant
    \citep{kandakoglu2024mcdm, chatterjee2018fuzzy} &
    No continuously updated MCDM weighting responsive to resource availability
    and project progress \\[0.4em]

    Skill demand forecasting &
    LSTM/TFT for labour market and HR demand; 7--25\% improvement over baselines
    \citep{macedo2022skills, lim2021tft} &
    No application to project-level, portfolio-wide skill pool demand across
    300+ concurrent CAPEX projects \\[0.4em]

    Hybrid AI / symbolic integration &
    Neurosymbolic taxonomy; ESCO-based skill extraction
    \citep{garcez2022neurosymbolic, gartner2024hype} &
    No hybrid AI system combining ESCO ontology, temporal demand forecasting,
    and prescriptive optimisation in a unified CAPEX framework \\
    \bottomrule
  \end{tabular}
\end{table}

The overarching research gap can be stated as follows:

\begin{quote}
  \textit{No existing work unifies (1)~strategic priority scoring via dynamic MCDM,
  (2)~AI-driven temporal modelling of specialised skill pool demand at portfolio scale, and
  (3)~prescriptive suspension-and-resumption optimisation into a single, operationally
  deployable framework for CAPEX portfolio management.}
\end{quote}

Each of the three components exists in isolation within its respective literature cluster.
The preemption literature provides the operational mechanism but lacks strategic intelligence
(Cluster~B). The MCDM literature provides the prioritisation logic but is static and
disconnected from scheduling (Cluster~E). The prescriptive analytics literature defines the
architectural blueprint but has not been applied to CAPEX portfolio decisions (Cluster~D).
The RCPSP literature addresses the computational problem at scale but ignores AI-driven
priority inputs (Cluster~C). The skill forecasting literature provides the temporal demand
models but has not been applied to portfolio-wide project staffing (Cluster~F).

Their integration, guided by the AI-4-SME Design--Build--Run methodology
\citep{peter2025ai4sme} and grounded in a hybrid AI architecture combining ESCO-based
symbolic knowledge with data-driven subsymbolic forecasting, constitutes the novel
contribution of this thesis. The three research sub-questions map directly onto the three
gap dimensions: SQ1 addresses skill demand modelling at portfolio scale; SQ2 addresses
dynamic MCDM weighting; and SQ3 evaluates the impact of the integrated suspension mechanism
on strategic portfolio performance.
