\chapter{Conclusion}
\label{chap:conclusion}

% TODO: Write conclusion (approx. 300–500 words)
% Summarize the research problem, the proposed approach, anticipated contributions,
% and directions for future research.

This proposal has presented a research design for developing and evaluating a hybrid AI
recommendation engine for CAPEX resource management. The central argument is that combining
ESCO-grounded knowledge graphs with data-driven demand forecasting models — structured according
to the AI-4-SME Design–Build–Run methodology — can substantially improve the anticipation of
skill bottlenecks and the quality of allocation recommendations compared to conventional
spreadsheet-based planning tools.

The four research questions address distinct aspects of this challenge: the design of the hybrid
AI architecture (RQ1), the selection of appropriate symbolic knowledge structures (RQ2), the
translation of detected bottlenecks into actionable recommendations (RQ3), and the empirical
evaluation of the resulting system (RQ4). Taken together, their answers will validate or refute
the thesis statement.

\textbf{Anticipated theoretical contributions:}
\begin{itemize}
  \item Design principles for knowledge-augmented forecasting systems in resource management.
  \item An empirically grounded evaluation of hybrid AI architectures in an applied
        organizational setting.
  \item Refinement of the AI-4-SME framework through application to a complex,
        multi-stakeholder CAPEX domain.
\end{itemize}

\textbf{Anticipated practical contributions:}
\begin{itemize}
  \item A prototype recommendation engine demonstrating how ESCO skill data and historical
        project staffing records can be combined to generate proactive workforce decisions.
  \item A replicable methodology for organizations seeking to implement similar systems
        within existing ERP/PPM infrastructure.
  \item Governance guidelines for deploying hybrid AI recommendation engines in compliance
        with GDPR and the EU AI Act.
\end{itemize}

\textbf{Future research:}
Future work could extend the evaluation to multiple organizations to test generalizability,
explore real-time integration with ERP systems, and investigate human-AI collaboration patterns
in the context of portfolio manager decision-making.
