\chapter{Theoretical Background}
\label{chap:background}

\section{CAPEX Portfolio Management}
\label{sec:capex_ppm}

\subsection{Definition and Characteristics}
% TODO

\subsection{Project Lifecycle Phases in CAPEX Contexts}
% TODO: initiation, planning, execution, monitoring, closeout

\subsection{Interdependency Structures in Large Portfolios}
% TODO: resource dependencies, schedule dependencies, technical dependencies

% ─────────────────────────────────────────────────────────────────────────────
\section{Resource Management in Project Portfolios}
\label{sec:resource_management}

\subsection{Skill-Based Resource Planning}
% TODO: skill taxonomies, competency frameworks

\subsection{Resource Bottlenecks: Definition and Impact}
% TODO: definition, types of bottlenecks, cost of late detection

\subsection{Current Planning Practices and Their Limitations}
% TODO: manual methods, spreadsheet-based tools, ERP resource modules

% ─────────────────────────────────────────────────────────────────────────────
\section{Forecasting Methods}
\label{sec:forecasting}

\subsection{Classical Time Series Methods}
% TODO: ARIMA, exponential smoothing

\subsection{Machine Learning for Demand Forecasting}
% TODO: gradient boosting, random forests, neural networks

\subsection{Deep Learning Approaches for Sequential Data}
% TODO: LSTM, Transformer-based models, temporal fusion transformers

% ─────────────────────────────────────────────────────────────────────────────
\section{Resource Allocation and Optimization}
\label{sec:optimization}

\subsection{Resource Allocation as an Optimization Problem}
% TODO: problem formulation, decision variables, constraints

\subsection{Mathematical Programming Approaches}
% TODO: LP, MIP, constraint programming

\subsection{Heuristic and Metaheuristic Methods}
% TODO: genetic algorithms, simulated annealing

% ─────────────────────────────────────────────────────────────────────────────
\section{The AI-4-SME Framework}
\label{sec:ai4sme_background}

The AI-4-SME Framework \citep{peter2025ai4sme}, developed at FHNW School of Business by Peter, Laurenzi, and Hinkelmann, provides a structured three-phase methodology — \textbf{Design, Build, Run} — for identifying and implementing AI solutions in enterprise contexts. It is adopted as the overarching methodological structure of this thesis.

\subsection{Design Phase: KIAs and DIAs}

A central concept of the framework is the distinction between \textit{Knowledge-Intensive Activities} (KIAs) and \textit{Data-Intensive Activities} (DIAs). KIAs require cognitive capabilities such as problem solving, analysis, creativity and decision making; DIAs primarily involve data analysis and transaction processing. In the CAPEX resource management context, skill-demand forecasting and bottleneck detection are KIAs, while allocation computation is a DIA.

\subsection{Build Phase: Hybrid AI Process Model}

The framework defines two development paths — a data-based path (ML with TensorFlow/PyTorch) and a knowledge-based path (knowledge graphs with Neo4j, GraphDB, Stardog, RDFox, metaphactory, or TopBraid Composer) — which can be combined into a hybrid AI solution. The interaction between both paths is a central feature of the Build phase.

\subsection{Run Phase: Governance and Integration}

Deployment addresses corporate governance (GDPR, EU AI Act), IT infrastructure decisions, API integration with ERP/CRM systems, and change management for human-AI collaboration.

\section{AI-Based Decision Support Systems}
\label{sec:dss}

\subsection{Recommendation Engines}
% TODO: collaborative filtering, content-based, hybrid approaches

\subsection{Explainability and Trust in AI Recommendations}
% TODO: SHAP, LIME, XAI requirements in enterprise settings
